---
title: "BAG R-Courses"
author: ["Patrick & Nicolas", "cynkra GmbH"]
date: "December 21, 2021"
output:
  cynkradown::cynkra_slides:
  # xaringan::moon_reader:
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: true
fontsize: 10pt
lang: english
font: frutiger
wide: false
colorlinks: false
logo: true
header-includes:
  - \usepackage{parskip}
---


# Efficient iteration

Two different approaches exist:

- for-loops
- apply-functions

**If done right**, both approaches are equally fast.
The myth that for-loops are slow is not true.

"for-loops" are a generic concept language-agnostic concept.
"apply functions" are mainly popular in R and not available in every language.

---

## The apply family

Package: {base}

**single list/vector**
- `tapply()`: Apply a function on a group of values
- `sapply()`: Apply a function over a list or vector (not typesafe!)
- `vapply()`: Apply a function over a list or vector

**multiple lists/vectors**
- `mapply()`: Apply a function to multiple list or arguments
- `lapply()`: Apply a function over a list and get a list in return
- `rapply()`: Recursive apply a fuction over a list of list

---

## {purrr} - tidyverse pkg for iteration

Pros:
- Typesafe: you always get a predefined type returned
  `map_chr()`, `map_int()`, `map_dbl()`, etc.
- Possibly simplified syntax
- Cat power (?)

Cons:
- Imports {rlang}
- Requires another pkg just for iteration

---

## purrr vs base apply

Create a matrix with 5 rows and 6 columns:

```{r }
mat <- matrix(rnorm(30), nrow = 5, ncol = 6)
mat <- tibble::as_tibble(mat) # tibble is an optimized dataframe
```

---

Now let´s sum the values of each column with `apply()`:

```{r}
apply(X = mtcars, MARGIN = 2, FUN = sum)
```

and with {purrr}:

```{r}
purrr::map_dbl(mtcars, sum)
```

---

## `vapply()` / `purrr:map_int()`

Count the number of characters of each name.

```{r}
names <- list("Antoine", "Mary", "Duncan", "Obalaya")
```

```{r}
vapply(names, nchar, FUN.VALUE = integer(1))
```

```{r}
purrr::map_int(names, nchar)
```

---

Exercise: Execute the same task but store the result in a list instead of a numeric vector

---

### Flexible argument placement

If the input is not the first argument of the applied function, one needs to be more explicit during the call specification.
Often, `function(x)` is used.
The "x" is an arbitray choice and could also be any other character (sequence).

```{r}
vapply(names, function(y) nchar(y), FUN.VALUE = integer(1))
```

```{r}
purrr::map_int(names, ~ nchar(.x))
```

{purrr} uses the `~` and `.x` syntax to "simplify" the iteration syntax.

---

## `tapply()`

`tapply()` is the 'group by and apply function' member of the 'apply-family'.

Return the mean of `Sepal.Length`, grouped by `Species`:

```{r}
tapply(iris$Sepal.Length, iris$Species, mean)
```

---


This can also be done **for all variables at once** combining `apply()` and `tapply()`.
However, the code line below is a bit advanced.
Let's break it down:

```{r}
apply(iris[, -5], MARGIN = 2, function(x) {
  tapply(x, iris$Species, mean)
})
```

1. In the `apply()` call, deselect the factor column which defines the grouping (`iris[, -5]`).
1. Apply `tapply()` to all columns (specified via ` MARGIN = 2`).

---

## `mapply()` / `purrr::map2*()`

Iterate over multiple vectors or lists at the same time.

Combine first and last names of the following vector and list:

```{r}
names_first <- c("Antoinette", "Mary", "Duncan")
names_last <- list("Miller", "Blidge", "Murdoc")
```

```{r}
mapply(paste, names_first, names_last, USE.NAMES = FALSE)
```

```{r}
purrr::map2_chr(names_first, names_last, ~ paste(.x, .y))
```

---

## rapply()

Recurisve apply on list of lists.
One use case would be an if/else replacement across all items of the nested lists.
To do so, we first define our own little function:

```{r}
myFun <- function(x) {
  return(paste(x, "!", sep = ""))
}
```

This function adds an "!" to each item.

```{r}
rl <- list(
  a = list(a1 = "Boo", b1 = "Eeek"),
  b = list(a2 = "Foo", b2 = list(a3 = "Hey"))
)
```

---

```{r}
rapply(l, function(x) myFun(x))
```

The result is an array.
To preserve the list structure you put in, use argument `how = "replace"`.

```{r}
rapply(l, function(x) myFun(x), how = "replace")
```

