---
title: "Tidyverse Intro II"
author: ["Antoine & Nicolas", "cynkra GmbH"]
date: "February 1st, 2022"
output:
  cynkradown::cynkra_slides:
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: true
fontsize: 10pt
lang: english
font: frutiger
wide: false
colorlinks: false
logo: true
header-includes:
  - \usepackage{parskip}
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
.remark-code {
    font-size: 12px;
}
.font17 {
    font-size: 17px;
}
.font14 {
    font-size: 14px;
}
</style>

# Introduction to {dplyr}

{dplyr} is the most important package of the tidyverse

We'll review its main functions, in this lesson we'll rename, subset (rows or columns) and sort the data.

* `rename()` to rename columns from a dataset
* `select()` to subset columns from a dataset
* `filter()` to subset rows of the data with a logical vector
* `slice()` to subset rows of the data with a numeric index
* `arrange()` to sort the data


???

{dplyr} contains many functions, they're all designed to do one simple action,
with a sensible default behavior. In this course we'll see how to use the basic
features of these functions.

---

# Dataset

We'll use the dataset `public_response` from the package {pixarfilms}

```{r, echo = FALSE}
options(tidyverse.quiet = TRUE)
```

```{r}
library(tidyverse)
library(pixarfilms)
public_response
```

---

# Rename columns with `rename()`

`rename()` keeps all columns and sets new names for specific columns

```{r}
# rename(<data frame>, <new name> = <old name>)
rename(public_response, movie = film)
```

???
`rename()` is one of the simplest {dplyr} verbs, the thing to remember is that
the new name is on the left and the old one on the right. 

---

# Rename columns with `rename()`

* Don't forget to assign the result!
* tidyverse functions never modify the input in place

```{r}
public_response_renamed <- rename(public_response, movie = film)
```

---

# Rename columns with `rename()`

* Use `(x <- ...)` to print and assign at the same time  

```{r}
(public_response_renamed <- rename(public_response, movie = film))
```

???
R either prints out the results, or saves them to a variable. If you want to do both, you can wrap the assignment in parentheses:

---

# Select columns with `select()`

What do you think these do ?

```{r, eval = FALSE}
select(public_response, rotten_tomatoes, cinema_score)
select(public_response, rotten_tomatoes:cinema_score)
select(public_response, -rotten_tomatoes)
select(public_response, -(rotten_tomatoes:cinema_score))
select(public_response, starts_with("c"))
select(public_response, contains("critic"))
select(public_response, where(is.numeric))
select(public_response, movie = film, rotten_tomatoes)
select(public_response, film, last_col())
select(public_response, cinema_score, everything())
```

???
`select()` is a very flexible function, thanks to a selection mechanism named
"tidy selection". Tidy selection can be used in several other functions, but
`select()` is the main one.

---

# Select columns with `select()`

```{r}
select(public_response, rotten_tomatoes, cinema_score)
```

---

# Select columns with `select()`

```{r}
select(public_response, rotten_tomatoes:cinema_score)
```

---

# Select columns with `select()`

```{r}
select(public_response, -rotten_tomatoes)
```

---

# Select columns with `select()`

```{r}
select(public_response, -(rotten_tomatoes:cinema_score))
```

---

# Select columns with `select()`

See also : `ends_with()`, `matches()`, `num_range()`

```{r}
select(public_response, starts_with("c"))
```

---

# Select columns with `select()`

See also : `ends_with()`, `matches()`, `num_range()`

```{r}
select(public_response, contains("critic"))
```


---

# Select columns with `select()`


```{r}
select(public_response, where(is.numeric))
```

---

# Select columns with `select()`


```{r}
select(public_response, movie = film, rotten_tomatoes)
```

---

# Select columns with `select()`

```{r}
select(public_response, film, last_col())
```

---

# Select columns with `select()`

```{r}
select(public_response, cinema_score, everything())
```

---

# Select columns with `select()`

What if I have column names stored in a vector ?

* use `all_of()` or `any_of()`

```{r}
cols <- c("film", "metacritic")
select(public_response, all_of(cols))
```

---

# Select columns with `select()`

They're almost the same!

* `all_of()` fails if some columns don't exist
* `any_of()` ignores columns that don't exist

```{r, error = TRUE}
cols <- c("film", "metacritic", "potatoe")
select(public_response, all_of(cols))
select(public_response, any_of(cols))
```

---

# Exercises

 *  Find several ways to select the 3 first columns
     
 *  What happens if you select a variable multiple ?
     
 *  Select all columns that contain underscores
   
 *  Use `all_of()` to select 2 columns of your choice
 
 *  Does the result of running the following code surprise you?  How do the
     select helpers deal with case by default? How can you change that default?
 
    `select(public_response, contains("R"))`


---

# Filter rows with `filter()`

* Filter a data frame with one or several filtering expressions
* Improved version of `subset()` base function

```{r, echo = FALSE}
options(tidyverse.quiet = TRUE)
```


```{r}
library(tidyverse)
library(pixarfilms)
# filter(<data frame>, <filtering expressions>)
filter(pixar_films, run_time == 100, film_rating == "G")
```

???
`filter()` allows you to subset observations based on their values. The first
argument is the name of the data frame. The second and subsequent arguments
are the expressions that filter the data frame. For example, we can select
all flights on January 1st with:

---

# Filter rows with `filter()`

<!-- TODO mention, this is a stats function, which is available by default? -->

* forgetting to load {dplyr} triggers surprising error
* base R has a different `filter()` function!

```{r, echo = FALSE, include= FALSE}
detach("package:dplyr")
```

```{r, error = TRUE}
filter(pixar_films, run_time == 100, film_rating == "G")
```

```{r, echo = FALSE, include= FALSE}
library(dplyr)
```

---

# Comparison operators

What is happening here ?

```{r, error = TRUE}
filter(pixar_films, run_time = 100)
```

---

# Comparison operators

* `=` is used to set arguments
* `==` is used to test equality
* Don't panic and read the error messages

```{r}
filter(pixar_films, run_time <= 92)
```

???
When you're starting out with R, the easiest mistake to make is to use `=`
instead of `==` when testing for equality. When this happens you'll get an
informative error:

---

# Comparison operators

* All comparison operators : `>`, `>=`, `<`, `<=`, `!=`, `==`

```{r}
filter(pixar_films, run_time <= 92)
```

???
To use filtering effectively, you have to know how to select the observations
that you want using the comparison operators. R provides the standard suite:
`>`, `>=`, `<`, `<=`, `!=` (not equal), and `==` (equal).

---

# Comparison operators

They work with character or date columns too!

```{r}
filter(pixar_films, film > "U")
subset(pixar_films, release_date < "2000-01-01")
```

---

# Comparison operators

What is happening here ?

```{r}
filter(pixar_films, number < 2)
```

???
the `number` column is stored as character! So our comparison is alhabetical

---

# Floating point numbers

What is happening here ?

```{r}
sqrt(2) ^ 2 == 2
1/49 * 49 == 1
near(c(sqrt(2) ^ 2, 1/49 * 49), c(2, 1))
```

???
There's another common problem you might encounter when using `==`: floating point numbers. These results might surprise you!
We eason with a mathematical model in mind, the computer knows
only decimal numbers and has to round them at some point. When this is an issue you can use `near()`

# Floating point numbers

stick with rounded numbers, or use `near()`

```{r}
near(sqrt(2) ^ 2,  2)
near(1 / 49 * 49, 1)
```

???
Computers use finite precision arithmetic (they obviously can't store an
infinite number of digits!) so remember that every number you see is an
approximation. Instead of relying on `==`, use `near()`:

near(sqrt(2) ^ 2,  2)
near(1 / 49 * 49, 1)

---

# Logical operators

* `&` : "and"
* `|` : "or"
* `!` : "not"

---

# Logical operators

* `&` : "and"
* `|` : "or"
* `!` : "not"

```{r}
# Same thing:
filter(pixar_films, run_time == 100, film_rating != "PG")

filter(pixar_films, run_time == 100 & film_rating != "PG")
```

???
Multiple arguments to `filter()` are combined with "and": every expression
must be true in order for a row to be included in the output. For other types
of combinations, you'll need to use Boolean operators yourself: `&` is "and",
`|` is "or", and `!` is "not".

---

# Logical operators

* `&` : "and"
* `|` : "or"
* `!` : "not"

```{r}
filter(pixar_films, film == "Toy Story" | run_time == "100")
```

---

# Logical operators

What is happening here ?

```{r, error = TRUE}
filter(pixar_films, film == "Toy Story" | "Toy Story 2")
filter(pixar_films, run_time == 81 | 82)
```

???
The order of operations doesn't work like English. You can't write
`filter(pixar_films, film == 81 | 82)`, which you might literally translate into
"finds all films with a run time of 81 or 82". Instead it finds
all months that equal `81 | 82`, an expression that evaluates to `TRUE`. In a
numeric context (like here), `TRUE` becomes one, so this finds all films with
a run_time of `1`, and there are none!

---

# Logical operators

* `"Toy Story" | "Toy Story 2"` fails
* ` 81 | 82` returns `TRUE`

```{r}
# this works
filter(pixar_films, film == "Toy Story" | film == "Toy Story 2")
# this is better
filter(pixar_films, film %in% c("Toy Story", "Toy Story 2"))
```

???
A useful short-hand for this problem is `x %in% y`. This will select every
row where `x` is one of the values in `y`. We could use it to rewrite the
code above:

---

# De Morgan's law

<!-- TODO maybe this is a bit overkill? -->

* simplify conditions when you can!
* `!(x & y)` is the same as `!x | !y`
* `!(x | y)` is the same as `!x & !y`

```{r}
filter(pixar_films, !(run_time < 93 | release_date > "2000-01-01"))
filter(pixar_films, run_time >= 93, release_date <= "2000-01-01")
```

???
Sometimes you can simplify complicated subsetting by remembering De Morgan's
law: `!(x & y)` is the same as `!x | !y`, and `!(x | y)` is the same as `!x &
!y`.

---

# Missing values

* R has a special value to say "I don't know" : `NA` (Not available) 
* They're a common occurence in datasets
* Watch out for "fake" NAs!

```{r}
tail(pixar_films)
```

???
It's common in "messy" data to find "NA" strings, it's often a symptom of
a failure at reading or writing the data properly.

---

# Missing values

What do these return ?

```{r, eval = FALSE}
NA > 5
10 == NA
NA + 10
NA / 2
NA == NA
```

???
One important feature of R that can make comparison tricky are missing
values, or `NA`s ("not availables"). `NA` represents an unknown value so
missing values are "contagious": almost any operation involving an unknown
value will also be unknown.
The last one might seem confusing, but if you have two unknown values, are they
equal ? you don't know!

---

# Missing values

```{r, eval = TRUE}
NA > 5
10 == NA
NA + 10
NA / 2
NA == NA
```

---

# Missing values

```{r}
# Let x be Mary's age. We don't know how old she is.
x <- NA

# Let y be John's age. We don't know how old he is.
y <- NA

# Are John and Mary the same age?
x == y
# We don't know!
```

---

# Missing values

* To determine if a value is missing, use `is.na()`
* We often use `!is.na()` in filter to remove missing values

```{r}
is.na(c(NA, "NA"))

filter(pixar_films, film == NA)

filter(pixar_films, is.na(film))
```

---

# Missing values

* `filter()` only includes rows where the condition is `TRUE`;
* It excludes both `FALSE` and `NA` values. 
* Preserve them explicitly if needed

```{r}
filter(pixar_films, run_time < 95 | is.na(run_time))
```

---

# Missing values

What do these return ?

```{r, eval = FALSE}
NA ^ 0
NA | TRUE
FALSE & NA
NA * 0
```

???

Missing values are often "contagious", it's not a rule however!

---

# Missing values

What do these return ?

```{r}
NA ^ 0
NA | TRUE
FALSE & NA
NA * 0
```

---

# Useful helpers

* On numeric: `between()`

```{r}
# same result!
filter(pixar_films, run_time >= 91, run_time <= 94)
filter(pixar_films, between(run_time, 91, 94))
```

<!-- TODO `%in%`, `near()`?-->

---

# Useful helpers

* On char: `str_starts()`, `str_ends()`, `str_detect()`
* from {stringr} package loaded with tidyverse

```{r}
filter(pixar_films, str_starts(film, "Toy"))

filter(pixar_films, str_detect(film, "ory"))
```

# Useful helpers

* On date: `year()`, `month()`, `day()`
* From the {lubridate} package NOT loaded with tidyverse

```{r}
library(lubridate, warn.conflicts = FALSE)

filter(pixar_films, year(release_date) == 1998)

filter(pixar_films, between(release_date, as_date("1998-01-01"), as_date("1998-12-31")))

filter(pixar_films, release_date >= "1998-01-01", release_date <= "1998-12-31")
```

<!-- TODO more fancy stuff like `filter_at(vars(-x, -y), all_vars(is.na(.)))`, just to show, there's more? -->

---

# `if_any()` and `if_all()`

What is happening here ?

```{r}
filter(pixar_films, if_any(c(film, run_time), is.na))

filter(pixar_films, if_all(c(film, run_time), is.na))
```

???
`if_any()` and `if_all()` allow advanced filter applying the same condition
on a set of variables, with `if_any()` we keep all the rows where the condition
is satisfied in at least one column, with `if_all()` we keep all the rows
where the condition is satisfied in all columns.

The first argument uses tidy selection, it means that we can use there everything
we've seen in `select()`

---

# Exercises

Find all films that

    1. Are rated "PG"
    2. Had a run time below 95 
    3. Had a rating of "N/A" or "Not Rated"
    4. Were released after and including year 2020 
    5. Have a missing name (`film` column) or `run_time`
    6. Were released after and including year 2020 with name not missing
    7. Are a first sequel (the name ends with "2")
    8. Were released between "2018-06-12" and  "2020-04-27" using `>=` and `<=`
    9. The same, using `between()` and `lubridate::as_date()`
    
---

# Subset rows by index with `slice()`

* `filter()` subsets rows with a logical vector,
* `slice()` subsets rows with a numeric vector
* The special function `n()` returns the index of the last row

```{r}
slice(pixar_films, 2:3)
```

---

# Subset rows by index with `slice()`

* `filter()` subsets rows with a logical vector,
* `slice()` subsets rows with a numeric vector
* The special function `n()` returns the index of the last row

```{r}
slice(pixar_films, 25:(n()-1))
```

---

# Subset rows by index with `slice()`

Other functions of the `slice()` family :

* `slice_head()` and `slice_tail()` select the first or last rows.
* `slice_sample()` randomly selects rows.
* `slice_min()` and `slice_max()` select rows with highest or lowest values of a variable.

```{r}
# n must be named!
slice_head(pixar_films, n = 2)
```

---

# Exercises

Using the `pixar_films` dataset :

 1.  Return the 2 first rows using `slice()` then using `slice_head()`

 2.  Return the 2 last rows using `slice()` then using `slice_tail()`

 3.  Return the rows of the 2 shortest movies using `slice_min()`

 4.  Return the rows of the 15% longest movies using `slice_max()` and the `prop` argument

---

# Arrange rows with `arrange()`

```{r}

arrange(pixar_films, run_time, release_date)
```

???
`arrange()` works similarly to `filter()` except that instead of selecting
rows, it changes their order. It takes a data frame and a set of column names
(or more complicated expressions) to order by. If you provide more than one
column name, each additional column will be used to break ties in the values
of preceding columns:

---

# Arrange rows with `arrange()`

Use desc to sort in descending order

```{r}
arrange(pixar_films, desc(film))
```

---

# Arrange rows with `arrange()`

Missing values are always sorted at the end

```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
arrange(df, desc(x))
```

---

# Arrange rows with `arrange()`

* We can arrange using expressions too
* sorting by logical -> **F**ALSE is before **T**RUE (like 0 is before 1)

```{r}
arrange(pixar_films, run_time < 110)
```

???
We generally arrange by existing columns but we can also use expressions.
Here we see that we didn't really sort by run time but all big run times are
brought on top

---

# Exercises

 1.  Arrange `pixar_films` by name but starting with missing values (Hint: sort using `!is.na()` and break the ties with `film`)

 2.  Arrange `pixar_films` to have the shortest film with a "PG" rating on top (Hint: use `film_rating != "PG"`)

 3.  Arrange `pixar_films` to have the longest film with a "PG" rating on top 

   
   
