---
title: "Avanced Tidyverse: advanced {dplyr}"
author: ["Antoine & Nicolas", "cynkra GmbH"]
date: "January 25, 2022"
output:
  cynkradown::cynkra_slides:
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: true
fontsize: 10pt
lang: english
font: frutiger
wide: false
colorlinks: false
logo: true
header-includes:
  - \usepackage{parskip}
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
.remark-code {
    font-size: 12px;
}
.font17 {
    font-size: 17px;
}
.font14 {
    font-size: 14px;
}
</style>

<!-- xaringan::inf_mr() -->

# {dplyr} refresher

```{r programming-with-dplyr-1}
library(dplyr, warn.conflicts = FALSE)
starwars
```

???
To start off we'll do a quick review on the main dplyr functions (often called
verbs). Then we'll dig in to learn some advanced features, especially useful
when programming.

We'll use the `starwars` dataset. It's loaded with the {dplyr} package and
contains a lot of cool information about starwars characters.

This dataset is a tibble, it's a special kind of data frame that prints nicer
and is easier to program with.

---

# {dplyr} refresher

`select()`

```{r programming-with-dplyr-2}
sw_selected <- select(starwars, name, height, homeworld, species)
sw_selected
```

???

The names function gives the column names of a tibble or data frame. 
Here we can see the names of the 13 columns of starwars dataset

The select function expects the columns we'd like to choose as unquoted 
arguments separated by commas. We can see the result of choosing a few columns here.

We'll see more advanced ways of selecting columns, through a feature called
"tidy selection"

---

# {dplyr} refresher

`filter()`

```{r programming-with-dplyr-3}
sw_filtered <- filter(sw_selected, homeworld %in% c("Alderaan", "Coruscant"))
sw_filtered
```

???

If we'd like to return only rows that match certain criteria, we use the filter function.

We'll see more advanced ways of filtering columns, through the features
`across()`, `if_any()`, `if_all()`

---

# {dplyr} refresher

`mutate()`

```{r programming-with-dplyr-4}
library(glue)
sw_mutated <- mutate(sw_filtered, height_m = height / 100)
sw_mutated
```

???

The mutate function can create a new column based either on combining multiple columns or on doing a calculation with a constant value. 

We'll see more advanced ways of mutating columns, through the features
`across()` , `rowwise()` and `c_across()`

---

# {dplyr} refresher

`group_by()` and `summarize()`

```{r programming-with-dplyr-5}
sw_grouped <- group_by(sw_mutated, species)
sw_summarized <- summarize(sw_grouped, avg_height_m = mean(height_m), .groups = "drop")
sw_summarized
```

???

Finally, we can use the group_by and summarize functions to aggregate values by groups.
Note the `.groups = "drop"` argument, it's not necessary here because each
`summarize()` call peels of one group, but in my opinion it's good practice to
always use it, unless we really want to keep groups, in which case
I recommend using an explicit `.groups = "keep"` arg.

We'll see more advanced ways of grouping and summarizing columns, building on what
we'll learn before.

---

# {dplyr} refresher

Full analysis

```{r programming-with-dplyr-6}
sw_selected <- select(starwars, name, height, homeworld, species)
sw_filtered <- filter(sw_selected, homeworld %in% c("Alderaan", "Coruscant"))
sw_mutated <- mutate(sw_filtered, height_m = height / 100)
sw_grouped <- group_by(sw_mutated, species)
sw_summarized <- summarize(sw_grouped, avg_height_m = mean(height_m), .groups = "drop")
sw_summarized
```

???
we see we created a lot of temp variables, they're only useful for the
next step, this can be simplified by using a single temporary variable

---

# {dplyr} refresher

Using a single temp variable

```{r programming-with-dplyr-7}
. <- starwars
. <- select(., name, height, homeworld, species)
. <- filter(., homeworld %in% c("Alderaan", "Coruscant"))
. <- mutate(., height_m = height / 100)
. <- group_by(., species)
. <- summarize(., avg_height_m = mean(height_m), .groups = "drop")
sw_summarized <- .
sw_summarized
```

---

# {dplyr} refresher

Using the pipe `%>%` (Ctrl + Shift + M), keeping dots explicit

```{r programming-with-dplyr-8}
sw_summarized <-
  starwars %>% 
  select(., name, height, homeworld, species) %>% 
  filter(., homeworld %in% c("Alderaan", "Coruscant")) %>% 
  mutate(., height_m = height / 100) %>% 
  group_by(., species) %>% 
  summarize(., avg_height_m = mean(height_m), .groups = "drop")
```

---

# {dplyr} refresher

Using the pipe `%>%` (Ctrl + Shift + M), with implicit dots

```{r programming-with-dplyr-9}
sw_summarized <-
  starwars %>% 
  select(name, height, homeworld, species) %>% 
  filter(homeworld %in% c("Alderaan", "Coruscant")) %>% 
  mutate(height_m = height / 100) %>% 
  group_by(species) %>% 
  summarize(avg_height_m = mean(height_m), .groups = "drop")
```

???
We often skip the dot argument, making it seem like we push the previous
result into the new function call. With some functions that don't have
their data argument first, we need to be explicit about the dot!

---

# Tidy selection

* {tidyselect} implements a dialect of R where operators make it easy
to select variables
* Used by several tidyverse functions, especially `dplyr::select()`


???

tidy selection is found in many places in dplyr and the rest of the tidyverse,
once you realize it many operations become much easier. We'll get familiars
with its features using `select()` and then will encounter it again later in the
course.

---

# Tidy selection

`:` for selecting a range of consecutive variables.

```{r programming-with-dplyr-11}
select(starwars[1:2,], name:eye_color)
```

???
You've probably used the colon operator `:` to define a sequence. With tidy 
selection we can use it to select a sequence of consecutive columns.

---

# Tidy selection

`!` for taking the complement of a set of variables.

```{r programming-with-dplyr-12}
select(starwars[1:2,], !name:eye_color)
```

???
Note that we can compose different tidy selection features, as we do above.

---

# Tidy selection

`!` for taking the complement of a set of variables.

```{r programming-with-dplyr-13}
select(starwars[1:2,], -name)
select(starwars[1:2,], !name)
```

???

`!` and `-` seem to be similar, do they do the same thing ?

---

# Tidy selection

`!` for taking the complement of a set of variables.

```{r programming-with-dplyr-14}
select(starwars[1:2,], -name, -height)
select(starwars[1:2,], !name, !height)
```

???
`!` is different from `-`, `-` is used to remove the columns that are specified,
while `!` takes the complement, the last calls takes the union of the 2 complements
so the first argument selects everything except names, and the second, everything
except `height`, repeated columns are ignored by tidy selection so the second
argument is only adding a `name` column at the end.

---


# Tidy selection

`c()` for combining selections

Useful for functions have a single tidyselect argument and we can't use `...`.

```{r programming-with-dplyr-15}
select(starwars[1:2,], c(name, eye_color))

select(starwars[1:2,], c(name:eye_color, gender:species))
```

???

Not so useful here since we can use several `...` args, but some 
functions have a single tidyselect argument and we can't use `...`.

---

# Tidy selection

`everything()` matches all variables.

```{r programming-with-dplyr-16}
select(starwars[1:2,], homeworld, everything())
```

???
You might wonder the point of selecting "everything", if you only select
`everything()` it is indeed a no-op, however as we saw, tidy selection prevents you
from selecting several times the same column, so `everything()` can be used
to reorder the columns.

It will find other uses with functions that apply a modification on a tidy selection
of columns, in that case "everything" will generally not be the default.

---

# Tidy selection

`last_col()` selects the last variable, possibly with an offset.

```{r programming-with-dplyr-17}
select(starwars[1:2,], homeworld, last_col())

select(starwars[1:2,], homeworld, last_col(c(1, 3)))
```

---

# Tidy selection

* `starts_with()` : Starts with a prefix.
* `ends_with()` : Ends with a suffix.
* `contains()`: Contains a literal string.

```{r programming-with-dplyr-18a}
select(starwars[1:2,], starts_with("s"))
```

???

These are quite easy to understand intuitively but be careful, they are case
insensitive by default! We recommend having all columns lower case much
as possible, in those case that won't make a difference

---

# Tidy selection

* `starts_with()` : Starts with a prefix.
* `ends_with()` : Ends with a suffix.
* `contains()`: Contains a literal string.

```{r programming-with-dplyr-18b}
select(starwars[1:2,], ends_with("color"))
```

---

# Tidy selection

* `starts_with()` : Starts with a prefix.
* `ends_with()` : Ends with a suffix.
* `contains()`: Contains a literal string.

```{r programming-with-dplyr-18c}
select(starwars[1:2,], contains("_"))
```

---

# Tidy selection

`matches()` : Matches a regular expression.

```{r programming-with-dplyr-19a}
select(starwars[1:2,], matches("^.a")) # string starts ('^') with any character ('.') then 'a'
```

???

`matches()` can be seen as a generalized version of `contains()`, 
regular expressions (aka regex) is a language of its own and anybody who manipulates
character strings often enough would gain from taking the time to learn them. 

---

# Tidy selection

Exercise!

How could you recreate these outputs without using `matches()` ?

```{r programming-with-dplyr-19}
select(starwars[1:2,], matches("^s"))      # '^' is the start of the string
select(starwars[1:2,], matches("color$"))  # '$' is the end of the string
```

???

However given the flexibility of the other features that tidy selection proposes,
there is often a way around using `matches()`

---

# Tidy selection

`num_range()` : Matches a numerical range like `x01`, `x02`, `x03`.

```{r programming-with-dplyr-21}
billboard_short <- tidyr::billboard[1:2, 1:15]
billboard_short

select(billboard_short, artist, num_range("wk", 3:8))
```

???

---

# Tidy selection

`&` and `|` for selecting the intersection or the union of two sets of variables.

```{r programming-with-dplyr-20}
select(starwars[1:2,], starts_with("s") | ends_with("color"))

select(starwars[1:2,], starts_with("s") & ends_with("color"))
```


---

# Tidy selection

`all_of()`: Matches variable names in a character vector. All names must be present, otherwise an out-of-bounds error is thrown.

```{r programming-with-dplyr-22a, error = TRUE}
cols <- c("name", "mass", "potatoe")
select(starwars[1:2,], all_of(cols))

cols <- c("name", "mass")
select(starwars[1:2,], all_of(cols))
```

???
Sometimes you'll have columns stored as character, `all_of()` and `any_of()`
are the most convenient ways to use them. 

---

# Tidy selection

`any_of()`: Same as all_of(), except that no error is thrown for names that don't exist.

```{r programming-with-dplyr-22b, error = TRUE}
cols <- c("name", "mass", "potatoe")
select(starwars[1:2,], any_of(cols))
```

---

# Tidy selection

`where()`: Applies a predicate function to all columns

```{r programming-with-dplyr-23}
select(starwars[1:2,], where(is.numeric))

select(starwars, where(~n_distinct(.x) < 10))
```

???

`where()` applies a predicate on the column's data, and not on the column names
as all the others we've seen.

It's often used to select variables according to type but more complexe predicates
are possible

---

# Tidy selection

Exercise!

Review all our tidy selection helpers and find 3 ways to select the columns 4 to 7 of the `starwars` dataset

---

# Tidy selection

* `relocate()` : used to move columns after of before other columns 
  (using `.after` or `.before` arguments)

```{r programming-with-dplyr-25}
relocate(starwars[1:2, 1:8], name, .after = height)
```

All arguments support tidy selection!

```{r programming-with-dplyr-26}
relocate(starwars[1:2, 1:8], c(name, height), .after = ends_with("color"))
```

???

This examples shows you the power of tidy selection, it's not reserved to
`select()` and has been implemented in all tidyverse functions where it makes sense.

---

# Advanced mutating

Use `.keep` to keep or drop columns depending on mutate expressions

```{r programming-with-dplyr-27}
# keep all, the default
mutate(starwars[1:2,], bmi = mass / height^2, .keep = "all")

# keep only computed columns and their sources
mutate(starwars[1:2, 1:8], bmi = mass / height^2, .keep = "used")
```

---

# Advanced mutating

Use `.keep` to keep or drop columns depending on mutate expressions

```{r programming-with-dplyr-27b}
# remove columns you don't need
mutate(starwars[1:2, 1:8], bmi = mass / height^2, .keep = "unused") 

# same as using `transmute()`
mutate(starwars[1:2, 1:8], bmi = mass / height^2, .keep = "none") 
```

???
This is a quite recent features that I have not seen much in the wild yet,
but it really nice compared to following `mutate()` by a `select()` call
in that it is much more compact and makes the intent explicit

---

# Advanced mutating

Use `across()` to apply a modification to columns using tidy selection

```{r programming-with-dplyr-28}
starwars[1:2, 1:5] %>%
  mutate(
    across(ends_with("color"), toupper),
    height_m = height / 100
  )
```

???

`across()` is also a recent addition, and a very important one because it renders
obsolete a full family of functions called "scoped variants" and suffixed with
"_at", "_if" or "_all".

---

# Advanced mutating

The default behavior when providing a named list is to use the name as a 
suffix for new columns.

```{r programming-with-dplyr-29}
mutate(starwars[1:2, 1:5], across(ends_with("color"), list(up = toupper)))
```

???
You might also provide several functions!

---

# Advanced mutating

If you need more control on the column names, the `.names` argument takes a
glue specification, with special values `.col` for the name of the column,
and `.fn` for the name of the function. 

```{r programming-with-dplyr-30}
mutate(starwars[1:2, 1:5], across(ends_with("color"), list(up = toupper), .names = "{.fn}{.fn}_{.col}"))
```

---

# Advanced mutating

Instead of functions we can use formulas

```{r programming-with-dplyr-31}
mutate(starwars[1:2, 1:5], across(where(is.numeric), ~ .x * 10))
```

---

# Advanced mutating

`c_across()` is a tidy selection aware version of `c()`, especially
useful with `rowwise()` 

```{r programming-with-dplyr-32}
# using `c()`
starwars[1:2, 1:5] %>% 
  rowwise() %>% 
  mutate(color_missing_rate = 
          c(hair_color, skin_color) %>% 
           is.na() %>% 
           mean()) %>%  
  ungroup()

# using `c_across()`
starwars[1:2, 1:5] %>% 
  rowwise() %>% 
  mutate(color_missing_rate = 
           c_across(ends_with("color")) %>% 
           is.na() %>% 
           mean()) %>%  
  ungroup()
```

---

# Advanced grouping and summarizing

`across()` works with `summarize()` too!

```{r programming-with-dplyr-33}
starwars %>% 
  group_by(sex) %>% 
  summarize(across(where(is.numeric), lst(min, max) , na.rm = TRUE))
```

---

# Advanced grouping and summarizing

* `group_by()` looks like a selecting verb, but it is a mutating verb!
 (in fact `group_by` works exactly like mutate, except that it sets groups)
* It means for instance we can `group_by(data, col1 + col2)`
* For this reason we cannot use tidy selection **directly** in `group_by()`

---

# Advanced grouping and summarizing

* However, we can use `across()` with a single argument!

```{r programming-with-dplyr-34}
starwars %>% 
  group_by(across(starts_with("se"))) %>% 
  summarize(across(where(is.numeric), lst(min, max) , na.rm = TRUE))
```

---

# Advanced grouping and summarizing

`count()` is another mutating verb that looks like a selecting verb.

Recognize the error to know when to use `across()`

```{r programming-with-dplyr-35, error = TRUE}
count(starwars, ends_with("color"))
```

---

# Advanced grouping and summarizing

`count()` is another mutating verb that looks like a selecting verb.

Recognize the error to know when to use `across()`

```{r programming-with-dplyr-35b}
count(starwars, across(ends_with("color")))
```

???

If you check the documentation of a function, you will see that arguments that
directly support tidy selection are documented as "<tidy-select>" (see `?select` or `?across`),
arguments documented as "<data-masking>" on the other hand will work with `across()`
so can leverage tidy selection indirectly as we see in above example.


---

# Advanced filtering

Can you guess what these do ?

```{r, eval = FALSE}
filter(starwars[1:8], if_any(ends_with("color"), ~ .x == "unknown"))

filter(starwars[1:8], if_all(ends_with("color"), ~ .x == "unknown"))
```

---

# Advanced filtering

* `if_all()` and `if_any()` apply the same predicate function to a selection of columns and combine the results into a single logical vector
* `if_any()` is `TRUE` when the predicate is `TRUE` for any of the selected columns
* `if_all()` is `TRUE` when the predicate is `TRUE` for all selected columns
* They'll be mostly useful in `filter()` calls

```{r}
filter(starwars[1:8], if_any(ends_with("color"), ~ .x == "unknown"))

filter(starwars[1:8], if_all(ends_with("color"), ~ .x == "unknown"))
```


---

# Advanced filtering

Exercise!

* Find all rows of `starwars` that have `NA`s in either `height`, `mass`, or `birth_year`
* Find all rows of `starwars` that have `NA`s in all `height`, `mass`, or `birth_year`

Hint : 

* This can be done using `if_any()` or `if_all()` or by testing columns separately
for missingness and using `|` or `&`. 
* Which is more compact ?
* Which is more readable ?

---

# <data-masking> vs <tidy-select>

The description of the arguments in the documentation tells you if you can
use tidy selection directly or across

```{r}
# arguments are described as "<tidy-select>", meaning variables refer to column positions
# we can use tidy selection directly
?select
?relocate
?across
?if_any

# arguments are described as "<data-masking>", meaning variables refer to values
# we can use `across()`
?mutate
?summarize
?group_by
?count
?filter
```

---

# Tunnelling data-variables

* We've learnt how to specify variables to tidyverse functions in many ways
* We still have to learn how to provide them through functions 

This doesn't work, why ?

```{r, error = TRUE}
mean_by <- function(data, by, var) {
  data %>%
    group_by(by) %>%
    summarise(avg = mean(var, na.rm = TRUE))
}

mean_by(starwars, gender, height)
```

???

The naive approach above doesn't work because `group_by()` and `summarize()`
look for the `by` and `var` variables in the data, whatever was provided
to `mean_by()` (`gender`, `height`) is not looked at

---

# Tunnelling data-variables

We could try to put into use what we have learnt about tidy selection, but we'd
need to provide strings to the function :

```{r, error = TRUE}
mean_by <- function(data, by, var) {
  data %>%
    group_by(across(all_of(by))) %>%
    summarise(across(all_of(var), mean, na.rm = TRUE, .names = "avg"))
}

mean_by(starwars, "gender", "height")
```

???

The naive approach above works but is not totally satisfying. The tidyverse
design guide recommends that variables be specified unquoted, and it'd be nice
to build functions that follow these principles

---

# Tunnelling data-variables

Tidyverse functions support a `{{` operator, called "curly curly" or
the "embrace" operator, we talk about "tunneling" a variable

Using it guarantees that the tidyverse functions sees the variable as it was
provided to the parent function

```{r, error = TRUE}
mean_by <- function(data, by, var) {
  data %>%
    group_by({{ by }}) %>%
    summarise(avg = mean({{ var }}))
}
mean_by(starwars, gender, height)
```

???

The naive approach above works but is not totally satisfying. The tidyverse
design guide recommends that variables be specified unquoted, and it'd be nice
to build functions that follow these principles

---

# Tunnelling data-variables

We can go even further and control the names of the arguments fed to the 
tidyverse functions. For this we need to use `:=` rather than `=`,
and use a glue a special glue syntax.

```{r}
mean_by <- function(data, by, var, prefix = "avg") {
  data %>%
    group_by({{ by }}) %>%
    summarise("{prefix}_{{ var }}" := mean({{ var }}, na.rm = TRUE))
}

mean_by(starwars, gender, height)
```

---

# Summary

* tidy selection : `:`, `!`, `starts_with()`, `where()`, `all_of()` ...
* Selecting functions recognize tidy selection directly : `select()`, `relocate()`, `across()`, `if_any()` ...
* Data masking functions recognize tidy selection through `across()`: `mutate()`, `summarize()`, `group_by()`, `filter()`
* `if_any()` and `if_all()` are useful in `filter`
* Use `{{` to feed naked variables to tidyverse functions just like they are fed
to their parent functions.
