---
title: "Tidyverse Intro III"
author: ["Antoine & Nicolas", "cynkra GmbH"]
date: "February 1st, 2022"
output:
  cynkradown::cynkra_slides:
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: true
fontsize: 10pt
lang: english
font: frutiger
wide: false
colorlinks: false
logo: true
header-includes:
  - \usepackage{parskip}
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
.remark-code {
    font-size: 12px;
}
.font17 {
    font-size: 17px;
}
.font14 {
    font-size: 14px;
}
</style>

# Introduction to {dplyr} (continued)

In this lesson we learn how to transform and aggregate our data,
with a detour to learn about the pipe operator

* `mutate()` to create new variables
* The pipe operator `%>%`
* `group_by()` and `summarize()` to compute aggregates

---

# Dataset

We'll use the datasets `pixar_films` and `public_response` from the package {pixarfilms}

```{r, echo = FALSE}
options(tidyverse.quiet = TRUE)
```

```{r}
library(tidyverse)
library(pixarfilms)
pixar_films
```

---

# Transform data with `mutate()`

Modify a variable

```{r}
mutate(pixar_films, run_time = run_time /60)
```

???
`dplyr::mutate()` is an improved version of `base::transform()`

---

# Transform data with `mutate()`

Create a new variable

```{r}
mutate(pixar_films, run_time_h = run_time /60)
```

---

# Transform data with `mutate()`

* Most of the time columns are vectors
* All the functions we use on vectors can be used here

```{r}
library(lubridate, warn.conflicts = FALSE)
mutate(pixar_films, label = paste0(film, " (", year(release_date), ")"))
```

???
Sometimes these columns are list or data frames but that's an advanced topic
There are many functions for creating new variables that you can use with
`mutate()`. 

The key property is that the function must be vectorised: it must
take a vector of values as input, return a vector with the same number of
values as output. 

A constant value is also accepted and will be recycled.

---

# Transform data with `mutate()`

* What is happening here ?

```{r}
mutate(
  public_response,
  cinema_score = NULL, # remove cinema_score column
  min_score = min(rotten_tomatoes, metacritic, critics_choice, na.rm = TRUE),
  max_score = max(rotten_tomatoes, metacritic, critics_choice, na.rm = TRUE)
)
```

???
Don't forget that we are playing with vector columns!

---

# Transform data with `mutate()`

* Luckily R has vectorized functions for most common tasks!

```{r}
mutate(
  public_response,
  cinema_score = NULL, # remove cinema_score column
  min_score = pmin(rotten_tomatoes, metacritic, critics_choice, na.rm = TRUE),
  max_score = pmax(rotten_tomatoes, metacritic, critics_choice, na.rm = TRUE)
)
```

---

# Transform data with `mutate()`

* We can reuse columns that we just created or modified

```{r}
mutate(
  public_response,
  cinema_score = NULL, # remove cinema_score column
  min_score = pmin(rotten_tomatoes, metacritic, critics_choice, na.rm = TRUE),
  max_score = pmax(rotten_tomatoes, metacritic, critics_choice, na.rm = TRUE),
  delta = max_score - min_score
)
```

---

# Transform data with `mutate()`

`transmute()` is a useful variant

```{r}
transmute(
  pixar_films,
  film, # note: no `=` here, just a col name
  release_year = year(release_date),
  label = paste0(film, " (", release_year, ")")
)
```

???
`transmute()` is a useful variant to avoid a redundant `select()` call

---

# Transform data with `mutate()`

* We can use both vectors and aggregated scalar values

```{r}
mutate(pixar_films, run_time_incr = run_time - median(run_time, na.rm = TRUE))
```

???
the median run time is 100 (Finding Nemo), it's a scalar value
We can subtract it from a vector

---

# Transform data with `mutate()`

Useful vectorised base operations in `mutate()` :

*   Arithmetic operators: `+`, `-`, `*`, `/`, `^`
*   Logical comparisons, `<`, `<=`, `>`, `>=`, `!=`
*   Logs: `log()`, `log2()`, `log10()`
*   Vectorized min and max : `pmin()`, `pmax()`
*   Cumulated operations : `cummin()`, `cummax()`, `cumsum()`, `cumprod()`

Useful aggregations : 

* `mean()`, `median()`, `sum()`, `min()`, `max()`, `sd()`

???
*   Arithmetic operators: `+`, `-`, `*`, `/`, `^`. These are all vectorised,
    using the so called "recycling rules". If one parameter is shorter than 
    the other, it will be automatically extended to be the same length. This 
    is most useful when one of the arguments is a single number: `air_time / 60`,
    `hours * 60 + minute`, etc.
    
    Arithmetic operators are also useful in conjunction with aggregate
    functions you'll learn about later. For example `y - mean(y)` 
    computes the difference from the mean.
*   Logical comparisons, `<`, `<=`, `>`, `>=`, `!=`, which you learned about
    earlier. If you're doing a complex sequence of logical operations it's 
    often a good idea to store the interim values in new variables so you can
    check that each step is working as expected.
*   Logs: `log()`, `log2()`, `log10()`. Logarithms are an incredibly useful
    transformation for dealing with data that ranges across multiple orders of
    magnitude. They also convert multiplicative relationships to additive.
*   Offsets: `lead()` and `lag()` allow you to refer to leading or lagging 
    values. This allows you to compute running differences (e.g. `x - lag(x)`) 
    or find when values change (`x != lag(x))`. They are most useful in 
    conjunction with `group_by()`, which you'll learn about shortly.
*   Cumulative and rolling aggregates: R provides functions for running sums,
    products, mins and maxes: `cumsum()`, `cumprod()`, `cummin()`, `cummax()`; 
    and dplyr provides `cummean()` for cumulative means. If you need rolling
    aggregates (i.e. a sum computed over a rolling window), try the RcppRoll
    package.

---

# Transform data with `mutate()`

The tidyverse has nice additions :

* date operations from {lubridate}
* string interpolation from {glue}
* string manipulation from {stringr} : `str_replace()`, `str_detect()`
* factor manipulation from {forcats} : `fct_relevel()`
* some more from {dplyr} : `cummean()`, `lag()`, `lead()`, `case_when()`, `if_else()`
* etc (there's a lot!)

???
We'll showcase a few of those can't go through everything, remember that
if you need to do something with dates, strings or factors, {lubridate},
{stringr} or {forcats} probably got you covered. Just typing `str_` or `fct_`
will trigger the autocomplete and you might find the function you need just by 
its name!

---

# Transform data with `mutate()`

Build strings in a readable way with `glue::glue()`

```{r}
library(glue)
# compare glue and paste0
#  mutate(pixar_films, label = paste0(film, " (", year(release_date), ")"))
mutate(pixar_films, label = glue("{film} ({year(release_date)})"))
```

---

# Transform data with `mutate()`

Do all sorts or string detection, extraction, substitution... with {stringr}.

```{r}
# replace space followed by number by empty string
mutate(pixar_films, franchise = str_replace(film, " \\d", ""))
```

???

For some you might have to learn a bit of regular expressions!

---

# Transform data with `mutate()`

`if_else` is useful to apply a conditional transformation or create
a binary category.

```{r}
mutate(pixar_films, run_time2 = if_else(run_time < 100, "short", "long"))
```

---

# Transform data with `mutate()`

`case_when()` is useful to avoid nested `if_else()` calls

```{r}
mutate(
  pixar_films,
  run_time2 = case_when(
  run_time < 95 ~ "short",
  run_time < 105 ~ "regular",
  TRUE ~ "long"
))
```

---

# Exercises

In `box_office` :

1.  Check that in `box_office`, `box_office_worldwide` is indeed the sum of
  the 2 other columns.
    
2.  Flag movies in `box_office` that did better in the US and Canada than in the rest of the
 world
    
3.  Use `if_else()` and `median()` to create a column `budget_type()` with
 values `"low budget"` and `"high budget"`
 
4.  Use `lag()` to create a column `days_since_previous_movie` in `pixar_films`
(The films are already sorted)

---

# Introducing the pipe operator

Recap : display the top 3 longest movies along with their release date

```{r}
films_selected <- select(pixar_films, film, release_date, run_time)
films_selected
```

---

# Introducing the pipe operator

Recap : display the top 3 longest movies along with their release date

```{r}
films_arranged <- arrange(films_selected, desc(run_time))
films_arranged
```

---

# Introducing the pipe operator

Recap : display the top 3 longest movies along with their release date

```{r}
films_filtered <- filter(films_arranged, !is.na(film))
films_filtered
```

---

# Introducing the pipe operator

Recap : display the top 3 longest movies along with their release date

```{r}
longest3 <- slice_head(films_filtered, n = 3)
longest3
```

---

# Introducing the pipe operator

It doesn't look too bad, but :

* We spent time giving names to all these variables we don't need further down.
* A reader might wonder if they'll be used again
* We communicate our intent ("longest3") only in the end

```{r}
films_selected <- select(pixar_films, film, release_date, run_time)
films_arranged <- arrange(films_selected, desc(run_time))
films_filtered <- filter(films_arranged, !is.na(film))
longest3 <- slice_head(films_filtered, n = 3)
longest3
```

???
This code is a little frustrating to write though because we have to give each 
intermediate data frame a name, even though we don't care about it. Naming 
things is hard, so this slows down our analysis. 

---

# Introducing the pipe operator

We could use a single temporary variable, let's try the dot `"."`

```{r}
. <- pixar_films
. <- select(., film, release_date, run_time)
. <- arrange(., desc(run_time))
. <- filter(., !is.na(film))
. <- slice_head(., n = 3)
longest3 <- .
longest3
```

---

# Introducing the pipe operator

* We still polluted our local environment with a temporary variableß
* We still communicate our intent ("longest3") only in the end
* The pipe operator `%>%` from {magrittr} solves those issues

```{r}
longest3 <-
  pixar_films %>% 
  select(., film, release_date, run_time) %>% 
  arrange(., desc(run_time)) %>% 
  filter(., !is.na(film)) %>% 
  slice_head(., n = 3)
  
longest3
```

???

The dot is a special variable when using the pipe, it means "what was computed before"

---

# Introducing the pipe operator

When a naked `"."` is not found in arguments, `%>%` assumes it should come first,
so we can simplify our call into:

```{r}
longest3 <-
  pixar_films %>% 
  select(film, release_date, run_time) %>% 
  arrange(desc(run_time)) %>% 
  filter(!is.na(film)) %>% 
  slice_head(n = 3)
  
longest3
```

???

It should feel as if we pushed the previous result into the next call.

This focuses on the transformations, not what's being transformed, which
makes the code easier to read. 
A good way to pronounce `%>%` when reading code is "then".
Our call can be read as a series of imperative
statements: select, then arrange, then filter, then slice. 


Behind the scenes, `x %>% f(y)` turns into `f(x, y)`, and `x %>% f(y) %>%
g(z)` turns into `g(f(x, y), z)` and so on. You can use the pipe to rewrite
multiple operations in a way that you can read left-to-right, top-to-bottom.

---

# Introducing the pipe operator

* In {tidyverse} the data argument is always first
* It is convenient for pipes!
* But we can use with base functions too, and not only on data frames

```{r}
1:10 %>% mean()
public_response %>% head(1)
public_response %>% lm(rotten_tomatoes ~ metacritic, .)
```

???
One important tidyverse principle is that functions have their data argument
first, this makes them work very well with the pipe, but you can use the pipe
with functions from the base packages or other packages too!

---

# Exercises

1. Replace "e" by "r" in the string "magritte" using first `str_replace()` then `sub()`,
without and with the pipe. What do you notice ?

2. With a single piped call, remove `cinema_score` from `public_response` and
`filter` to keep only movies that did `90` or better on all other review websites

---

# Aggregate with `group_by()` and `summarize()`

`summarize()` collapses a data frame to a single row:

```{r}
summarize(public_response, mean_rt_score = mean(rotten_tomatoes, na.rm = TRUE))
```

---

# Aggregate with `group_by()` and `summarize()`

* Most often we use `summarize()` with `group_by()`
* `group_by()` makes the date "grouped" so `summarize()` and other functions
can make use of groups.

```{r}
# group_by(<data frame>, <grouping variables>)
# summarize(<data frame>, <summarizing expressions>)
public_response %>%
  group_by(cinema_score) %>%
  summarize(
    mean_rt_score = mean(rotten_tomatoes, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
```

???
`summarize()` is not mostly useful when we pair it with `group_by()`.
This changes the unit of analysis from the complete dataset to individual
groups. Then, when you use the dplyr verbs on a grouped data frame they'll be
automatically applied "by group". For example, if we applied exactly the same
code to a data frame grouped by cinema score, we get the average rotten tomatoes score per
cinema score.

---

# Aggregate with `group_by()` and `summarize()`

`count()` is a helpful shortcut for `data %>% group_by(...) %>% summarize(n = n())`

```{r}
public_response %>% 
  group_by(cinema_score) %>% 
  summarize(n = n(), .groups = "drop")

public_response %>% 
  count(cinema_score)
```

---

# Aggregate with `group_by()` and `summarize()`

Exercises!

Look at the `academy` dataset

1. Count the nominations for every movie (Hint:you'll need `filter()`)
2. A nomination is 1 point, winning an award is 2 points, compute every movie's
 score and show the top 3 (Hint: create a `points` column using `mutate()` and `case_when()`)
 
---

# Using `group_by()` with other verbs

Groups are most useful with `summarize()` but other verbs can use them too!

Don't forget to `ungroup()` !

```{r}
public_response %>% 
  group_by(cinema_score) %>% 
  mutate(median_metacritic = median(metacritic, na.rm = TRUE)) %>% 
  ungroup()
```

---

# Using `group_by()` with other verbs

Groups are most useful with `summarize()` but other verbs can use them too!

```{r}
public_response %>% 
  group_by(cinema_score) %>% 
  filter(metacritic > median(metacritic, na.rm = TRUE)) %>% 
  ungroup()
```

---

# Using `group_by()` with other verbs

Groups are most useful with `summarize()` but other verbs can use them too!

```{r}
public_response %>% 
  group_by(cinema_score) %>% 
  filter(metacritic > median(metacritic, na.rm = TRUE)) %>% 
  ungroup()
```
